High-order dynamic neural correlations reflect naturalistic processing in humans

### 2019-07-29 paper/main.tex
updated figure 2 size to be less crazy
updated brain plot code and reworded introduction

\begin{abstract}
  Our thoughts arise from coordinated activity patterns across our
  brain.  We examined high-order dynamic correlations in functional
  neuroimaging data collect as human participants listened to
  different auditory stimuli varying in cognitive richness, along with
  an additional resting state condition.  Our approach combines a
  kernel-based method for estimating the dynamic functional
  correlations that are similar (within task) across participants,
  along with a dimensionality reduction approach that enables us to
  efficiently compute high-order correlations in the data.  We trained
  classifiers to decode the precise time, relative to the start of the
  stimulus, when a given neural pattern was recorded.  We trained
  these classifiers using the neural activity timeseries, first-order
  dynamic correlations, and higher-order correlations (up to
  tenth-order correlations), and asked which types of features led to
  the highest decoding accuracy.  We found that second-order
  correlations consistently yielded the highest decoding accuracy in
  all of the listening conditions of the experiment, whereas
  first-order correlations yielded the highest decoding accuracy at
  rest.
\end{abstract}

@@

\section*{Introduction}
The dynamics of the observable universe are meaningful in three
respects.  First, the behaviors of the \textit{atomic units} that
exhibit those dynamics are highly interrelated.  The actions of one
unit typically have implications for one or more other units.  In
other words, there is non-trivial \textit{correlational structure}
defining how different units interact with and relate to each other.
Second, that correlational structure is \textit{hierarchical} in the
sense that it exists on many spatiotemporal scales.  The way one group
of units interacts may relate to how another group of units interact,
and the interactions between those groups may exhibit some rich
structure.  Third, the structure at each level of this correlational
hierarchy changes from moment to moment, reflecting the ``behavior''
of the full system.

These three properties (rich correlations, hierarchical organization,
and dynamics) are major hallmarks of many complex systems.  For
example, within a single cell, the cellular components interact at
many spatiotemporal scales, and those interactions change according to
what that single cell is doing.  Within a single human brain, the
individual neurons interact within each brain structure, and the
structures interact to form complex networks.  The interactions at
each scale vary according to the functions our brains are carrying
out.  And within social groups, interactions at different scales
(e.g., between individuals, family units, communities, etc.) vary over
time according to changing goals and external constraints.

Although many systems exhibit rich dynamic correlations at many
scales, a major challenge to studying such patterns is that typically
neither the correlations nor the hierarchical organizations of those
correlations may be directly observed.  Rather, these fundamental
properties must be inferred indirectly by examining the observable
parts of the system-- e.g., the behaviors of the individual atomic
units of that system.  In the \textit{Methods} section, we propose a
series of mathematical operations that may be used to recover dynamic
correlations at a range of scales (i.e., orders of interaction).  In
the \textit{Results} section, we demonstrate how our approach may be
applied to multi-dimensional timeseries data: a synthetic dataset
where the underlying dynamic correlations are known (we use this
dataset to validate our approach), and a neuroimaging dataset
comprising data collected as participants listened to a
story~\citep{SimoEtal16}.  In different experimental conditions in the
neuroimaging study, participants listened to altered versions of the
story that varied in cognitive richness: the intact story (fully
engaging), a scrambled version of the story where the paragraphs were
presented in a randomized order (moderately engaging), a second
scrambled condition where the words were presented in a random order
(minimally engaging), and a ``rest'' condition where the participants
did not listen to any version of the story (control condition).  We
use the neuroimaging dataset to examine how higher-order structure in
brain data varies as a function of the cognitive richness of the
stimulus.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/methods_fig}
  \caption{\textbf{Computing higher order correlations. } Dynamic
    correlations
  are computed then approximated to the same size as original data
  using dimensionality reduction.
  This process is repeated, and can be used to compute up to any
  arbitrary order with compuations scaling linearly (as opposed to exponentially) with order. }
  \label{fig:methods_fig}
\end{figure}

>>>>

\begin{abstract}
  Our thoughts arise from coordinated activity patterns across our
  brain.  We examined high-order dynamic correlations in functional
  neuroimaging data collect as human participants listened to
  different auditory stimuli varying in cognitive richness, along with
  an additional resting state condition.  Our approach combines a
  kernel-based method for estimating the dynamic functional
  correlations that are similar (within task) across participants,
  along with a dimensionality reduction approach that enables us to
  efficiently compute high-order correlations in the data.  We trained
  classifiers to decode the precise time, relative to the start of the
  stimulus, when a given neural pattern was recorded.  We trained
  these classifiers using the neural activity timeseries, first-order
  dynamic correlations, and higher-order correlations (up to
  tenth-order correlations), and asked which types of features led to
  the highest decoding accuracy.  We found that higher-order
  correlations consistently yielded higher decoding accuracy for
  the most cogntively rich condition of the experiment, whereas
  first-order correlations or raw data yielded the highest decoding accuracy at
   for the less cogntively rich conditions as well as rest.
\end{abstract}

@@

\section*{Introduction}
To understand the neuronal computations that support cognition, we must understand the cooperative dynamics of populations of neurons.  These populations of neurons interact within each brain structure, and the structures interact to form complex and dynamic networks.  These interactions, at each scale, vary according to the functions our brains are carrying out.

In the last several decades, advances in functional magnetic resonance image (fMRI) analyses have evolved to better characterize these interactions of brain structures.  Resting-state (RS) fMRI has shed light on the spatiotemporal organization of spontaneous brain activity, with rich and complex network organization.  Functional connectivity (FC) has further characterized the task dependency of these networks.  However, recent work has shown that FC fluctuates over time (Chang and Glover, 2010), and the assumed stationarity of these analyses may be too simple to capture the dynamic nature of brain activity.

Similarly, multivariate activity patterns have given an advance over univariate activity patterns because they allow relative contributions of voxels to be combined.  Furthermore, recent work show that dynamic correlations between multivariate voxel patterns can add an additional boost (Manning, 2018).

Following this trajectory of analyses, one possible next step would be to better characterize the dynamic patterns of connectivity.  To this end, we wished to understand if higher order cognitive processing is mediated by dynamic higher-order correlations.

However, a major challenge to studying such patterns is that typically neither the correlations nor the hierarchical organizations of those correlations may be directly observed.  Rather, these fundamental properties must be inferred indirectly by examining the observable parts of the system-- e.g., the behaviors of the individual units of that system.  In the \textit{Methods} section, we propose a series of mathematical operations that may be used to recover dynamic correlations at a range of scales (i.e., orders of interaction).  In the \textit{Results} section, we demonstrate how our approach may be applied to multi-dimensional timeseries data: a synthetic dataset where the underlying dynamic correlations are known (we use this dataset to validate our approach), and a neuroimaging dataset comprising data collected as participants listened to a story~\citep{SimoEtal16}.  In different experimental conditions in the neuroimaging study, participants listened to altered versions of the story that varied in cognitive richness: the intact story (fully engaging), a scrambled version of the story where the paragraphs were presented in a randomized order (moderately engaging), a second scrambled condition where the words were presented in a random order (minimally engaging), and a ``rest'' condition where the participants did not listen to any version of the story (control condition).  We use the neuroimaging dataset to examine how higher-order structure in brain data varies as a function of the cognitive richness of the stimulus.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{figs/methods_fig}
  \caption{\textbf{Computing higher order correlations. } Dynamic
    correlations
  are computed then approximated to the same size as original data
  using dimensionality reduction.
  This process is repeated, and can be used to compute up to any
  arbitrary order with compuations scaling linearly (as opposed to exponentially) with order. }
  \label{fig:methods_fig}
\end{figure}
